---
title: "Class 11: Structural Bioinformatics 1"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Structural Bioinformatics

## Section 1: Introduction to the RCSB Protein Data Bank (PDB)

Q1: Download a CSV file from the PDB site (accessible from “Analyze” -> “PDB Statistics” >
“by Experimental Method and Molecular Type”. Move this CSV file into your RStudio project
and determine the percentage of structures solved by X-Ray and Electron Microscopy. Also can
you determine what proportion of structures are protein? 
```{r}
data <- read.csv("Data Export Summary.csv")
data
```

```{r}

ans <- data$Total/sum(data$Total) *100
names(ans) <- data$Experimental.Method
round(ans,2)
```

Also can you determine what proportion of structures are protein?
```{r}
areProt <- sum(data$Proteins)/sum(data$Total)
round(areProt, 2)
```

## Section 3: Bio3d
```{r}
library(bio3d)
pdb <- read.pdb("1hsg")
pdb
```

```{r}
str(pdb)


```

Atom selection
```{r}
ca.inds <- atom.select(pdb, "calpha")

```

Write new PDB of only protein
```{r}
prot <- atom.select(pdb, "protein", value =  TRUE)
write.pdb(prot, file = "1hsg_protein.pdb")
```


Write new PDB of only ligand
```{r}
ligand <- atom.select(pdb, "ligand", value = TRUE)
write.pdb(ligand, file = "1hsg_ligand.pdb")
```

