---
title: "Class 12: Structural Bioinformatics"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Prepare protein structure for docking

Download 1HSG PDB structure and then produce separate protein-only and ligand-only PDB files
```{r}
library(bio3d)

get.pdb("1hsg")

```

Produce "1hsg_protein.pdb" and "1hsg_ligand.pdb" files
```{r}
pdb <- read.pdb("1hsg.pdb") 
pdb
```

```{r}
ligand <- atom.select(pdb, "ligand", value = TRUE)
write.pdb(ligand, file = "1hsg_ligand.pdb")

protein <- atom.select(pdb, "protein", value = TRUE)
write.pdb(protein, file = "1hsg_protein.pdb")
```

## Preparing docking results for vmd

```{r}
res <- read.pdb("all.pdbqt", multi=TRUE) 
write.pdb(res, "results.pdb")
```

## RMSD: Quantitative assessment of docking results

```{r}
res <- read.pdb("all.pdbqt", multi=TRUE) 
ori <- read.pdb("ligand.pdbqt")
rmsd(ori, res)
```

## Normal Mode Analysis (NMA)
Normal mode analysis (NMA) is one of the major simulation techniques used to probe large- scale motions in biomolecules. Typical application is for the prediction of functional motions in proteins.
Normal mode analysis (NMA) of a single protein structure can be carried out by providing a PDB object to the function nma(). In the code below we first load the Bio3D package and then download an example structure of hen egg white lysozyme (PDB id 1hel) with the function read.pdb(). Finally the function nma() is used perform the normal mode calculation:

```{r}
pdb <- read.pdb("1HEL")
modes <- nma(pdb) 
plot(modes, sse=pdb)
```

To visualize the normal mode predicted structural variations we can use the function mktrj() to generate a trajectory PDB file by interpolating along a given normal mode:
```{r}
 # Visualize NMA results
mktrj(modes, mode=7, file="nma_7.pdb")
```

